Species motif participation provides unique information about species risk of extinction

### 2023-06-06 manuscript/JAE_reply.tex
Updates from Overleaf

\section*{Reply to Reviewer \#1} [[Done?]]

@@

                \textbf{R:} The sentence the reviewer refers to is ``How disturbances such as population decline or extinction lead to the loss of species via  direct and indirect effects (i.e., secondary extinctions) has long been a vibrant area of research (Curtsdotter et al., 2011a; Dunne & Williams, 2009; Eklöf \& Ebenman, 2006 Santos et al., 2021).'' We can understand the reviewer´s confusion, but in fact species extinctions are both the disturbance and the response.
                An initial extinction of a species (i.e. the primary extinction) can, since species are connected to each other via  consumer--resource interactions, cause other species to go extinct (i.e. secondary extinctions). We have now clarified this on line XX:

@@

                \textbf{R:} To avoid introducing extra terminology, we have replaced all instances of `global' and `local' with `large-scale' and `small-scale' respectively. We hope it is now easier to understand motifs as a meso-scale descriptor of network structure. See line XX, quoted below, for a short description. We have also, as described above, explained how motif participation captures information about a focal species' interaction partners, and thereby the potential for indirect interactions. We do not refer to second-step indirect interactions in order to avoid confusion with that term.

                \begin{quotation}
                Therefore, a species' motif participation profile captures the indirect interactions the species are affected by, thereby capturing important structures that are missed by global (network level) and local (species level) properties.  As such, motif participation may complement degree and trophic level to more fully predict species' extinction risk.
                \end{quotation}

@@

            \item Line 92-95: The authors bring up many conceptual topics in their introduction, but perhaps the most relevant topic: network motif structures, has been almost entirely overlooked in terms of providing background information to the reader about which motifs are most common, how they influence energy transfer, and what encourages the establishment of the different motif structures across food webs. The authors mention that they will be investigating food chain, omnivory, apparent and direct competition motifs only briefly at the end of the introduction. From my perspective, the authors are missing key details about what is already known about how these different structural configurations of sub-webs can influence stability and consumer persistence. For example, networks with a high degree of omnivory may be more stable than simple food chains, but this depends on the strength of the omnivorous interactions among trophic guilds. [[Did they take this from our discussion, the point they found contradictory?]] It would be helpful to provide more insight into the gaps in knowledge surrounding the prevalence of these motifs in food webs more generally, and why this would help food web researchers understand consumer persistence and how it is related to resilience/stability since this is where I see the greatest amount of novelty from this article.

                \textbf{R:} Every study we are aware of that relates motifs to stability either models one or more motifs in isolation or examines the effect of motif profiles (or elements thereof, e.g., frequency of omnivory) on the whole network. Neither one is exactly applicable to the effect of motif participation on per-species extinction risk. The exception is an earlier study by some authors of the present work, which we mention but do not wish to over-promote by making this paper a keystone of the introduction. For these reasons, together with constraints of length, we have been judicious in our selection of additional material on previous motif-stability relationships. Particularly for omnivory, the existing literature paints a picture of context-dependence and variable effects rather than a clear general trend. Reviewing this material would be more suited to a general review (perhaps with supplemental modelling to fill gaps between tests of different types of stability, disturbance, etc.) than to a paragraph in our introduction. We hope that the refocusing and revision of the introduction provides enough clarity to compensate for this necessarily cursory overview.

@@

                \textbf{R:} We do think that our statement here is valid. Topological models for simulating species extinctions are only taking the food web structure in to account, and a secondary extinction will occur when a consumer lost all its resources. So, in comparison to topological models simulating the biomass (or similar) dynamics between many species will require both more parameters and simulation time. If we compare to for example the ATN-model requires parameters describing each species body mass and several parameters for defining for example attack rates and handling times. Similarly, models like EcoPath with EcoSim, is limited to smaller food webs with well-understood interaction strengths and initial population sizes and as such require more parameters which even in simple cases are more complex than topological models. However, these more complex models can model more complex mechanisms than what topological models could.

                %We are not aware of any dynamical model that is fast over a large food web - certainly this was not the case when some of the authors conducted dynamical simulations of webs with 50-100 species. [[Anyone have a better answer? From the focus on major energy flows I think it's EcoSim they're thinking of...]]

@@

                \textbf{R:} Please note that we are using a Bayesian network modelling framework and not Bayesian statistics. All statistical models are quite standard linear models or general linear models. The Bayesian network method has been used and validated in previous papers (see references in the main text) and has not been substantially modified for this paper. The novelty in our work lies in the relationships between motifs and persistence, not in the network modelling \emph{per se}.

@@

                \begin{quotation}
                We generated simulated networks using the niche model, which has been shown to recreate the structure of empirical networks well~\citep{Williams2000,Stouffer2007}.
                To capture a range of network architectures similar to those in well-studied empirical networks~\citep{Dunne2002,Dunne2002a}, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
                All networks were generated using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}.   
                For full details and a comparison to empirical networks, see \emph{Appendix S2}.
                \end{quotation}

@@

                \textbf{R:} Proportion of role is the proportion of the species' motif role made up by the particular motif in each panel. This misunderstanding seems linked to a general difficulty in understanding precisely how we define motif roles. To help remedy this, we have now added motif role vectors to Figure 1 to make it explicit how counts of motifs are translated into vectors of normalised motif frequencies. These frequencies are equivalent to the proportion of a species' total motif participation made up by each motif. We also now use the perminology ``proportion of motif in role'' and ``proportion of motif in network profile'' in the axis labels of Figs. 2, 3, and 5 for consistency. We have also thoroughly revised the results section to emphasize our main findings and place supplemental results in the proper context.

@@

                \textbf{R:} Using the Bayesian network framework, we simply cannot include top-down effects on persistence. However, in a parallel study (started around the same time but by chance published earlier) some of the authors of the present study did simulate both top-down and bottom-up effects in a dynamical model. The dynamical model has its own caveats (in general, dynamical models are sensative to parameter choice and, due to the time and computational resources required, a full sensitivity analysis could not be performed). However, [[something comparing the results between papers - somewhat similar?]]

                It may also reassure the Reviewer to know that in the initial studies introducing the Bayesian network framework~\citep{}, Bayesian networks captured XX\% of the extinctions identified by dynamical models run on the same set of networks. As we use the same modelling framework as in those studies, we can reasonably expect similar results.

@@

                    There are limitations to the approach presented here.
                    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
                    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
                    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
                    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2021_inprep}.
                    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
                    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.

@@

    \bibliographystyle{ecollett} 
    \bibliography{manual} % Do not abbreviate journal titles, papers with >

>>>>

\section*{Reply to Reviewer \#1}

@@

                \textbf{R:} An initial extinction of a species (i.e. the primary extinction) can, since species are connected to each other via  consumer--resource interactions, cause other species to go extinct (i.e. secondary extinctions). We have now clarified this on line XX:

@@

                \textbf{R:} To avoid introducing extra terminology, we have replaced all instances of `global' and `local' with `large-scale' and `small-scale' respectively. We hope it is now easier to understand motifs as a meso-scale descriptor of network structure. See line XX, quoted below, for a short description of motifs and definition of meso-scale structure. We have also, as described above, explained how motif participation captures information about a focal species' interaction partners, and thereby the potential for indirect interactions. We do not refer to second-step indirect interactions in order to avoid confusion with that term.

                \begin{quotation}
                    Motifs are unique patterns of interacting species that correspond to different sets of direct and indirect interactions, providing a holistic description of meso-scale (i.e., intermediate between overall and fine-scale) network structure~\citep{Stouffer2007,Stouffer2012}.
                \end{quotation}

@@

            \item Line 92-95: The authors bring up many conceptual topics in their introduction, but perhaps the most relevant topic: network motif structures, has been almost entirely overlooked in terms of providing background information to the reader about which motifs are most common, how they influence energy transfer, and what encourages the establishment of the different motif structures across food webs. The authors mention that they will be investigating food chain, omnivory, apparent and direct competition motifs only briefly at the end of the introduction. From my perspective, the authors are missing key details about what is already known about how these different structural configurations of sub-webs can influence stability and consumer persistence. For example, networks with a high degree of omnivory may be more stable than simple food chains, but this depends on the strength of the omnivorous interactions among trophic guilds. It would be helpful to provide more insight into the gaps in knowledge surrounding the prevalence of these motifs in food webs more generally, and why this would help food web researchers understand consumer persistence and how it is related to resilience/stability since this is where I see the greatest amount of novelty from this article.

                \textbf{R:} We have expanded our description of existing knowledge about motifs but have kept this section relatively brief for several reasons. First, nearly every study we are aware of that relates motifs to stability either models one or more motifs in isolation or examines the effect of motif profiles (or elements thereof, e.g., frequency of omnivory) on the whole network. Neither one is exactly applicable to the effect of motif participation on per-species extinction risk. One exception is an earlier study by some authors of the present work, which we mention but do not wish to over-promote by making this paper a keystone of the introduction. Second, in order to place more of the methods in the main text while complying with length restrictions we must keep the introduction brief.
                We hope that the refocusing and revision of the introduction provides enough clarity to compensate for this necessarily cursory overview.

@@

                \textbf{R:} We do think that our statement here is valid. In large food webs, there are very many parameters affecting the population dynamics of interacting species. While in empirical systems it may be possible to ignore many of these parameters and focus on a small set of `key' parameters as you suggest, as in models like EcoPath with EcoSim, these are limited to smaller food webs with well-understood interaction strengths and initial population sizes.
                In simulated networks, selection of such key parameters would either be highly arbitrary (and potentially strongly affecting model results) or require an extensive (and computationally costly) sensitivity analysis. 
                On the other hand, the ATN-model requires parameters describing each species' body mass and several parameters for defining for example attack rates and handling times. 
                This is many more than either topological or Bayesian network models and leads to a much longer computation time.
                Note that we do not say that dynamical models are not useful (indeed, we have used them ourselves in different contexts), only that the computation time required for large networks prevents us from exploring multiple disturbance scenarios.

@@

                \textbf{R:} Please note that we are using a Bayesian network modelling framework and not Bayesian statistics. All statistical models are  standard linear models or general linear models. The Bayesian network method has been used and validated in previous papers (see references in the main text) and has not been substantially modified for this paper. The novelty in our work lies in the relationships between motifs and persistence, not in the network modelling \emph{per se}.

@@

                \begin{quotation}
                We generated simulated networks using the niche model, which has been shown to recreate the structure of empirical networks well~\citep{Williams2000,Stouffer2007}.
                To capture a range of network architectures similar to those in well-studied empirical networks~\citep{Dunne2002,Dunne2002e}, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
                All networks were generated using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}.   
                For full details and a comparison to empirical networks, see \emph{Appendix S2}.
                \end{quotation}

@@

                \textbf{R:} Proportion of role is the proportion of the species' motif role made up by the particular motif in each panel. This misunderstanding seems linked to a general difficulty in understanding precisely how we define motif roles. To help remedy this, we have now added motif role vectors to Figure 1 to make it explicit how counts of motifs are translated into vectors of normalised motif frequencies. These frequencies are equivalent to the proportion of a species' total motif participation made up by each motif. We also now use the terminology ``proportion of motif in role'' and ``proportion of motif in network profile'' in the axis labels of Figs. 2, 3, and 5 for consistency. We have also thoroughly revised the results section to emphasize our main findings and place supplemental results in the proper context.

@@

                \textbf{R:} Using the Bayesian network framework, we simply cannot include top-down effects on persistence. However, in a parallel study (started around the same time but by chance published earlier) some of the authors of the present study did simulate both top-down and bottom-up effects in a dynamical model. The dynamical model has its own caveats (in general, dynamical models are sensitive to parameter choice and, due to the time and computational resources required, a full sensitivity analysis could not be performed). However, we note that a dynamical analysis on the same set of networks (albeit with a different disturbance and persistence metric) found substantial consistency in motif-persistence relationships.

                It may also reassure the Reviewer to know that in the initial studies introducing the Bayesian network framework~\citep{Eklof2013}, Bayesian networks captured the majority of the extinctions identified by dynamical models run on the same set of networks. As we use the same modelling framework as in those studies, we can reasonably expect similar results.

@@

                    There are limitations to the approach presented here.
                    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
                    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
                    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
                    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2022Oikos}.
                    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
                    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.

@@

    \bibliographystyle{jae} 
    \bibliography{anna_bib_new} % Do not abbreviate journal titles, papers with >
